\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{\mbox{H.\arabic{table}}}%
\renewcommand{\thefigure}{\mbox{H--\arabic{figure}}}%
%

\chapter{Java and UML Resources}

\markboth{{\color{cyan}APPENDIX\,H\,\,$\bullet$\,\,}Java and UML Resources}
{{\color{cyan}APPENDIX\,H\,\,$\bullet$\,\,}Java and UML Resources}

\section*{Reference Books}
\begin{BL}

\item  David Flanagan, {\it Java in a Nutshell Edition 5}, 5th ed.,
O'Reilly and Associates 2005. Part of the O'Reilly Java series, this
book provides a concise desktop reference to Java and the API.

\item  James Gosling, Bill Joy, and Guy Steele, {\it The Java
Language Specification}, 3d ed., Addison-Wesley, 2005. This book,
which is part of Addison-Wesley's Java Series, provides a detailed
description of the Java language.  An online version is available at

\WWW
\begin{jjjlisting}
\begin{lstlisting}[commentstyle=\color{black}]
http://java.sun.com/docs/books/jls
\end{lstlisting}
\end{jjjlisting}

\item  Martin Fowler, {\it UML Distilled}, 3d ed., 
Addison-Wesley, 2003. This book, which is part of Addison-Wesley's
Object Technology Series, provides a concise introduction to UML.
\end{BL}

\section*{Online References}
\begin{BL} 

\item  {\tt http://www.omg.org/} contains good information on UML.

\item  {\tt http://java.sun.com/j2se} is one of Sun
Microsystems' Java Web sites.  From this page you can find links
to downloads of JDK, API specifications, and documentation
on all of Java, including Swing, AWT, and new features of
Java 5.0.

\item  {\tt http://java.sun.com/docs/codeconv/} provides a
description of coding conventions suggested by the {\it Java Language
Specification} and followed by the Java programming community.
(These are summarized in Appendix~A.)

\item  {\tt http://java.sun.com/tutorial} provides an online
Java tutorial.

\item  {\tt http://www.JARS.com} provides reviews and 
ratings of the best Java applets.

\item {\tt http://www.java-news-center.org/} is a clearinghouse for
Java news, programming examples, debugging tips, and many other useful
resources.

\end{BL}
%
